---
title: "Reproducible Research: Peer Assessment 1"
output: 
  md_document:
      variant: markdown_github
date: "27 de agosto de 2018"
keep_md: true
author: "Eduardo Pacheco"

---
```{r}

knitr::opts_chunk$set(echo = TRUE)

```

# Loading and preprocessing the data

```{r}

#first we will download the data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = "PA1.zip")
unzip("PA1.zip")

#now we have activity.csv. Lets see the content
activity <- read.csv("activity.csv",sep=",",  na.strings="NA")
head(activity)
tail(activity)
summary(activity)
str(activity)

# I prefer to use dataframes so...
 database <- data.frame()
 database <- rbind(database, activity)
 
 #now lets process the data : remove missing data and to date the dates
 database$date <- as.Date(database$date)
 database2 <- subset(database, !is.na(database$steps))

```

## What is mean total number of steps taken per day?


```{r}
daysum <- tapply(database2$steps, database2$date, sum, na.rm=TRUE, simplify=T)
daysum <- daysum[!is.na(daysum)]

hist(x=daysum,
     col="purple",
     breaks=25,
     xlab="total steps ( day)" ,
     ylab="Freq",
     main="dist of daily steps ( without missing)")
```

Now calculate the mean and median of steps/day

```{r}
mean(daysum)

median(daysum)
```


## What is the average daily activity pattern?
I would create a time series plot  of the 5 minutes interval (x) and the average number of steps(y)

```{r}
database3 <- tapply(database2$steps, database2$interval, mean, na.rm=TRUE, simplify=T)
database4 <- data.frame(interval=as.integer(names(database3)), avg=database3)

with(database4,
     plot(interval,
          avg,
          type="l",
          xlab="5 min intervals",
          ylab="avg steps in the interval across all days"))
```

Which avg( 5 minutoes interval) contains the maximum number of steps?

```{r}
max_steps <- max(database4$avg)
database4[database4$avg == max_steps, ]
```


## Imputing missing values
Now we need to calculate the total number of missing values in dataset
```{r}
sum(is.na(database$steps))
```

our database started with 2304 nulls ( NA's).

we will include the mean in NA's
```{r}
df_impute <- database
ndx <- is.na(df_impute$steps)
int_avg <- tapply(database2$steps, database2$interval, mean, na.rm=TRUE, simplify=T)
df_impute$steps[ndx] <- int_avg[as.character(df_impute$interval[ndx])]
```
now we will use the historgram of total steps /day and calculate the median total number of steps/day
```{r}
new_dailysum <- tapply(df_impute$steps, df_impute$date, sum, na.rm=TRUE, simplify=T)

hist(x=new_dailysum,
     col="purple",
     breaks=25,
     xlab="day steps",
     ylab="freq",
     main="The distribution of daily total (with missing data imputed)")
```

```{r}
mean(new_dailysum)
median(new_dailysum)
```


## Are there differences in activity patterns between weekdays and weekends?

We need to create two factor variables. One for weekend and other for weekday or whathever the word for this in english. 

```{r}
is_weekday <- function(d) {
    wd <- weekdays(d)
    ifelse (wd == "saturday" | wd == "sunday"| wd =="sat" | wd =='sun', "weekend", "weekday")
}

wx<- sapply(df_impute$date, is_weekday)
df_impute$wk <- as.factor(wx)
head(df_impute)



```


```{r}
wk_df <- aggregate(steps ~ wk+interval, data=df_impute, FUN=mean)

library(lattice)
xyplot(steps ~ interval | factor(wk),
       layout = c(1, 2),
       xlab="Interval",
       ylab="Number of steps",
       type="l",
       lty=1,
           data=wk_df)
```

